Implementation of a minimal branch-decomposition algorithm for planar graphs.

### 2024-08-12 paper/main.tex
Update paper

	A edge $e$ where $\phi_G(e) = \{u,v\}$, is a \textit{self-loop}, if $u = v$.

	A graph $G$ is \textit{loop-less}, if no edge $e \in \mathbb{E}(G)$ is a self-loop.

@@

	A \textit{walk} of a graph $G$ is a list of edges, such that consecutive edges in the list are incident to each other.

	An \textit{$s,t$-walk} is a walk, such that the first edge is incident to the vertex $s$ and such that the last edge is incident to the vertex $t$.

@@

	An $s,t$-walk is \textit{closed}, if $s=t$.

	A \textit{path} of a graph $G$, is a walk such that only consecutive edges of the walk are incident.

	A \textit{cycle} of a graph $G$, is a path but where the first and the last edge are incident to each other.

@@

	A \textit{bijection} is a relation between two sets such that each element of either set is paired with exactly one element of the other set.

@@

	\begin{definition}
		(\textit{Contraction})

		A contraction is a function that given a multi-graph $G$ and pair of vertices $u,v \in V(G)$ such that $\{u,v\} \in E(G)$, then for all edges $e \in \mathbb{E}(G)$ if $\phi_G(e) = \{u,v\}$ then removes $e$ else if $\phi_G(e) = \{v,w\}$ then $\phi_G(e) = \{u,w\}$, and finally returns the resulting graph.
	\end{definition}

@@

	\begin{definition}
		(\textit{Medial Graph})

@@

	An \textit{internal vertex} $v$ of a tree $T$, is a vertex $v \in V(T)$ of degree at least 2.

	Let the \textit{leaf set} of a tree $T$, denoted $L(T)$, be the set of all vertices of $T$ that are also leaves.

	A \textit{Branch Decomposition} $(B_G, \delta_G)$ of a graph $G$ consists of firstly, a tree $B_G$ where every internal vertex of $B_G$ has degree 3; in other words, $B_G$ is an unrooted binary tree, and secondly a bijection $\delta_G$ between $\mathbb{E}(G)$ and $L(B_G)$.

@@

	A \textit{Carving Decomposition} $(C_G, \lambda_G)$ of a graph $G$ consists of firstly, a tree unrooted binary tree $C_G$ and secondly a bijection $\lambda_G$ between $V(G)$ and $L(C_G)$.

@@

\section{Description of the problem}

@@

	\begin{definition}
		\textsc{The Planar Minimal Branch Decomposition Problem}

@@

	Here are some informal definitions to unpack the aforementioned properties of graphs.

	\begin{definition}\label{def:mbdp}
		\textsc{The Minimal Branch Decomposition Problem}

		Input: Given a simple connected graph $G$.

	The algorithm described in this paper solves \textsc{The Planar Minimal Branch Decomposition Problem}, which can be computed in polynomial time.

	The width of a minimal branch decomposition of $G$ is called the branch width of $G$.

\section{The algorithm}
	This section describes the algorithm given by Seymour and Thomas\cite{ST93} by identifying a set of practical problems and subproblems and how they relate.

@@

	Considering a graph $G$, you can compute a minimal branch decomposition $(B_G, \delta_G)$ of $G$ from a minimal carving decomposition $(C_{G^\times}, \lambda_{G^\times})$ of the medial graph $G^\times$ of $G$. Do this by replacing the vertices in leaves of the carving decomposition with edges, using the mapping between edges and vertices from computing the medial graph;

	Therefore problem \ref{def:mbdp} break down into problems \ref{problem:mcd-to-mbd}, \ref{problem:medial} and \ref{problem:mcd}.

@@

	\begin{problem}\label{problem:medial}
		Given a graph $G$, output a medial graph and a bijectional mapping between medial nodes and vertex pairs.
	\end{problem}

	\begin{problem}\label{problem:mcd}
		Given a plane graph $M$ and function to compute the carving width of a graph, output a minimal carving decomposition of $M$.
	\end{problem}

	Implementing a function to solve \ref{problem:mcd-to-mbd} is described in \ref{impl:mcd-to-mbd}.

	To solve \ref{problem:medial} is a matter of following the definition.

	Computing a medial graph is described in \ref{impl:medial}.

@@

	Doing a series of edge contractions (contraction of all edges incident to a pair of vertices) on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices can be assembled into a minimal carving decomposition of $M$.

	I will defer describing exactly how to assemble a minimal carving decomposition to \ref{impl:mcd}.

@@

	First consider problem \ref{problem:contraction}.

	Informally; a contraction is merging two vertices into one and letting incident edges connect to the new vertex. The contraction in this paper differs a bit from conventional definitions of edge- and vertex-contractions, as other definitions might result in self-loops when contracting one of multiple parallel edges.

	Computing a medial graph is described in \ref{impl:medial}.

	Now consider problem \ref{problem:cw}.

	The rat-catching algorithm decides $cw(M) \geq k$ with $k$ being in the positive integers. This is a monotonic boolean space, so you can perform a binary search to find the smallest $k$ where $cw(M) \geq k$ is true.

	The rat-catching algorithm can be described as a game of two players, the rat and rat-catcher. Considering a graph $M$, the edges of a face can be thought of as walls of a room and vertices as the corners of some rooms. The rat moves from corner to corner along the walls and the rat-catcher moves from room to room through some wall. The rat-catcher can force the rat away from some walls by making noise. A round of this game is played with some noise level $k$. The rat-catcher wins if they can force the rat to be in some wall of the room that they are in, with noise level $k$, and the rat wins if there is a strategy whereby the rat can escape indefinitely.

@@

	So assuming $\Delta(M) < k$ the game is played.

@@

	\begin{definition}\label{def:face-quiet}
		When the rat-catcher is on some room $f$, then edge $e$ is noisy iff. there is a closed walk of length scrictly less than $k$ containing $f^*$ and $e^*$ in the dual $M^*$.
	\end{definition}

@@

\section{The implementation}

	For the upcoming problems one needs to deal with parallel edges and be able to tell them apart, therefore the implementation uses a data structure that encapsulates an adjacency list of edges and a map from unique edge ids its vertexpair.

	\pyfile{graph.py}{0}{100}

	\subsection{Computing a minimal branch decomposition}\label{impl:mcd-to-mbd}

	\subsection{Computing a medial graph}\label{impl:medial}

		I assume that the input graph $G$ is a rotation system of a planar graph; an adjacency list such that the neighborhoods are given in clockwise ordering according to some plane embedding of $G$.

		Given this format, any two consecutive edges $e_1$ and $e_2$ in some face of $G$ are therefore consecutive vertices in the neighborhood of the vertex that $e_1$ and $e_2$ share.

		The implementation adds all medial links around some vertex for each vertex in $G$.

		The clockwise ordering of neighborhoods of $G$ becomes counterclockwise ordering of neighborhoods of the medial $M$. The medial graph of a plane graph is 4-regular \ref{}. From the perspective of some medial node $v$, in some single iteration of the loop on line 14, two links are added to the neighborhood of $v$ in counterclockwise ordering, and later the two other links are added to the neighborhood of $v$ also in counterclockwise ordering.

		\pyfile{medial_graph.py}{4}{22}

	\subsection{Computing a minimal carving decomposition}\label{impl:mcd}

		Solving problem \ref{problem:mcd}.

		The sequence of contracted edges is found and reassembled into a minimal carving decomposition.

		The "contraction" function returns a new unique vertex ID, therefore by saving which vertex is a contraction of which vertex pair in the "edges" dictionary, constructing the decomposition is then a matter of recursively expanding any vertices that were a result of a contraction into a tuple of the vertex pair that is was composed of. Repeating this until all only vertices of $M$ remain gives a carving decomposition in Newick-like nested tuple format.

	\subsection{Contraction}\label{impl:contraction}

		As this contraction is a contraction of ALL edges connecting a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the rotation system could be difficult, but luckily in this context, it is irrelevant.

		For a contraction of vertices $a$ and $b$, I have chosen to create a new vertex ID $c$ instead of reusing $a$ or $b$ as this later makes assembling the carving decomposition easier.

		This is where telling apart parallel edges, which the Graph class allows, becomes very useful. Inferring where to stitch together the neighborhoods to preserve the ordering, just from a normal adjacency list, becomes a way harder problem.

	\subsection{Carving width and the rat cathing algorithm}\label{impl:cw}

		Solving problem \ref{problem:cw}

@@

	\subsection{Quiet subgraph}\label{impl:quiet}

@@

		No other path of the implementation needs the assumption that the dual is planar, therefore no clockwise or counterclockwise ordering of the neighborhoods of the adjacency list is needed.

@@

		% https://en.wikipedia.org/wiki/Rotation_system

@@

\newpage
\section{Appendix}
	\pyfile{branch_decomposition.py}{0}{1000}
	\pyfile{branch_width_brute_force.py}{0}{1000}	
	\pyfile{branch_width.py}{0}{1000}
	\pyfile{carving_decomposition.py}{0}{1000}
	\pyfile{carving_width_brute_force.py}{0}{1000}
	\pyfile{carving_width.py}{0}{1000}
	\pyfile{contraction.py}{0}{1000}
	\pyfile{dual_graph.py}{0}{1000}
	\pyfile{Graph.py}{0}{1000}
	\pyfile{medial_graph.py}{0}{1000}
	\pyfile{parse_graph.py}{0}{1000}
	\pyfile{parse_newick.py}{0}{1000}

>>>>

	A \textit{self-loop}, is an edge $e$ where $\phi_G(e) = \{u,v\}$ and $u = v$.

	A graph $G$ is \textit{loop-less}, if no edge $e \in \mathbb{E}(G)$ is a self-loop. The edge with label 7, in the graph above, is an example of a self-loop.

@@

	$$\Delta(G) = 4$$

	A \textit{walk} of a graph $G$ is a list $[v_0,e_1,v_1,...,e_k,v_k]$ where $v_0,v_1,...,v_k \in V(G)$ and for $1 \leq i \leq k$, $\phi_{G}(e_i) = \{v_{i-1}, v_i\}$.

	A walk of $G$ $$[4,7,4,4,5,5,3,5,5,6,3]$$

@@

	An \textit{s,t-walk} is a walk where $s = v_0$ and $t = v_k$.

	An $s,t$-walk is \textit{closed}, if $s = t$.

	A \textit{path} of a graph $G$, is a walk such that no vertex is repeated in the list.

	A \textit{cycle} of a graph $G$, is an $s,t$-walk such that no vertex is repeated in the list except $s=t$.

@@

	A \textit{bijection} (or \textit{one-to-one correspondence}) is a relation between two sets such that each element of either set is paired with exactly one element of the other set.

@@

	A \textit{rotation system} is an encoding of a graph, in particular, it is an adjacency list such that the neighborhood, around any vertex, is in clockwise ordering according to some plane embedding of $G$.

	\begin{definition}\label{def:contraction}
		(\textit{Contraction})

		A contraction is a function that given a multi-graph $G$ and pair of distinct vertices $u,v \in V(G)$ such that $\{u,v\} \in E(G)$, then for all edges $e \in \mathbb{E}(G)$ if $\phi_G(e) = \{u,v\}$ then removes $e$ else if $\phi_G(e) = \{v,w\}$ then $\phi_G(e) = \{u,w\}$, and finally returns the resulting graph.
	\end{definition}

@@

	\begin{definition}\label{def:medial}
		(\textit{Medial Graph})

@@

	Let the \textit{leaf set} of a tree $T$, denoted $L(T)$, be the subset of vertices $L(T) \subseteq V(T)$ that are also leaves of $T$.

	An \textit{internal vertex} $v$ of a tree $T$, is a vertex $v \in V(T) - L(T)$ that is not a leaf. An internal vertex therefore has degree at least 2.

	An \textit{unrooted binary tree} $T$, is a tree where every internal vertex has degree 3.

	A \textit{Branch Decomposition} $(B_G, \delta_G)$ of a simple graph $G$ consists of firstly, an unrooted binary tree $B_G$ and secondly a bijection $\delta_G$ between $\mathbb{E}(G)$ and $L(B_G)$.

	\begin{center}
		\input{images/branch.tex}
	\end{center}

@@

	A \textit{Carving Decomposition} $(C_G, \lambda_G)$ of a simple graph $G$ consists of firstly, an unrooted binary tree $C_G$ and secondly a bijection $\lambda_G$ between $V(G)$ and $L(C_G)$.

@@

\section{Overview of the algorithm}

	Given a simple connected planar graph $G$, the algorithm computes a minimal branch decomposition of $G$.

	To compute a minimal branch decomposition of $G$, the algorithm first computes the medial graph $G^\times$ of $G$, secondly computes a minimal carving decomposition of $G^\times$, and finally computes a minimal branch decomposition of $G$ from the minimal carving decomposition of $G^\times$.

	To compute the minimal carving decomposition of $G^\times$, the algorithm uses a contraction algorithm that finds a series of contractions that does not increase the carving width and then assembles the series of contractions into a minimal carving decomposition of $G^\times$.

	To compute the carving width of $G$, the algorithm uses the rat-catching algorithm. By an analogy, the rat-catching algorithm can be described by a game of two players, the rat and rat-catcher, who take turns moving around on $G$. The game is played to determine, if the rat-catcher can corner the rat by making noise and thereby scaring away the rat from some subgraph, the size of which is dependent on $k$ or if the rat can escape indefinitely. Larger $k$ makes the rat-catching game easier for the rat-catcher. The smallest $k$ where the rat can escape indefinitely is the carving width of $G$.

	To compute a noisy subgraph of $G$ given a noise level $k$ and a rat-catchers position, the algorithm considers the dual graph of $G$ and finds the set of dual edges reachable by a closed walk of length at most $k-1$, the dual of these edges is the noisy subgraph.

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{./images/overview.png}
		\caption{Overview of the algorithm}
		\label{fig:overview}
	\end{figure}

\section{Data Structures and implementation considerations}

	For many subroutines the algorithm needs to deal with parallel edges and be able to tell them apart, therefore the implementation encodes a graph as an adjacency list of edges and a map from unique edge ids its vertexpair.

	\pyfile{graph.py}{0}{33}

\section{The algorithm}

@@

	\begin{definition}\label{def:mbdp}
		\textsc{The Planar Minimal Branch Decomposition Problem}

@@

	The algorithm described in this paper solves \textsc{The Planar Minimal Branch Decomposition Problem} in polynomial time.

	This section describes the algorithm given by Seymour and Thomas\cite{ST93} by identifying a set of practical problems and subproblems and how they relate.

@@

	Considering a graph $G$, you can compute a minimal branch decomposition $(B_G, \delta_G)$ of $G$ from a minimal carving decomposition $(C_{G^\times}, \lambda_{G^\times})$ of the medial graph $G^\times$ of $G$.

	Therefore problem \ref{def:mbdp} breaks down into subproblems \ref{problem:mcd-to-mbd}, \ref{problem:medial} and \ref{problem:mcd}.

	\begin{problem}\label{problem:medial}
		Given a plane graph $G$, output a medial graph $G^\times$, along with a bijectional mapping between medial nodes $V(G^\times)$ and vertex pairs $E(G)$.
	\end{problem}

	\begin{problem}\label{problem:mcd}
		Given a plane graph $M$, output a minimal carving decomposition of $M$.
	\end{problem}

@@

	With subroutines for all three problems, the implementation obtains a branch decomposition like so.

	\pyfile{branch_decomposition.py}{0}{100000}

	Here are three sections, one for each subproblem.

	\subsection{Medial graph}

		Recall the definition \ref{def:medial}.

		I assume that the input graph $G$ is a rotation system. Given this format, any two consecutive edges $w$ and $v$ in some face of $G$ are therefore consecutive vertices in the neighborhood of the vertex $a$ that $w$ and $v$ share.

		\begin{figure}[H]
			\centering
			\includegraphics[width=\textwidth]{./images/medial-implementation-4-reg-argument.png}
			\caption{Two iterations of the algorithm computing a medial rotation system}
			\label{fig:overview}
		\end{figure}

		A medial graph is 4-regular; every node has degree 4. From the perspective of some medial node $v$, in some single iteration of the loop on line 14, two links are added to the neighborhood of $v$ in counterclockwise ordering, and later the two other links are added to the neighborhood of $v$ also in counterclockwise ordering. Therefore the medial graph $M$ is a rotation system.

		\pyfile{medial_graph.py}{4}{24}

	\subsection{Minimal Branch Decomposition from Minimal Carving Decomposition}

	\subsection{Minimal Carving Decomposition}

@@

		By doing a series of edge contractions on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices, can be assembled into a minimal carving decomposition of $M$.

		The "contraction" function returns a new unique vertex ID, therefore by keeping which vertex is a contraction of which vertex pair in the "edges" dictionary, constructing the decomposition is then a matter of recursively expanding any vertices that were a result of a contraction into a tuple of the vertex pair that is was composed of. Repeating this until all only vertices of $M$ remain gives a carving decomposition in Newick-like nested tuple format.

@@

		\subsubsection{Contraction}

			Recall the definition \ref{def:contraction}. As this contraction is a contraction of ALL edges connecting a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the rotation system could be difficult, but in this context, it is irrelevant.

			For a contraction of vertices $a$ and $b$, I have chosen to create a new vertex ID $c$, instead of reusing $a$ or $b$, as this makes the implementation, for assembling the carving decomposition, simpler.

			This is where telling apart parallel edges, which the Graph class allows, becomes very useful. Inferring where to stitch together the neighborhoods to preserve the ordering, solely from an adjacency list, is way less practical.

		\subsubsection{Carving Width}

			Solving problem \ref{problem:cw}.

			The rat-catching algorithm decides $cw(M) \geq k$ with $k \in \mathbb{N}^+$.

			The rat-catching algorithm can be described as a game of two players, the rat and rat-catcher. Considering a graph $M$, the edges of a face can be thought of as walls of a room and vertices as the corners of some rooms. The rat moves from corner to corner along the walls and the rat-catcher moves from room to room through some wall. The rat-catcher can force the rat away from some walls by making noise. A round of this game is played with some noise level $k$. The rat-catcher wins the round if they can force the rat to be in some wall of the room that they are in, and the rat wins the round if there is a strategy whereby the rat can escape indefinitely.

@@

			So if $\Delta(M) < k$, then the game is played to determine an outcome, otherwise the rat is said to win.

			The outcomes of the rounds for $k=\{1,2,3,...\}$ is a monotonic boolean space, so you can perform a binary or linear search to find the smallest $k$ where $cw(M) \geq k$ is true.

			\pyfile{carving_width.py}{0}{8}
			\pyfile{carving_width.py}{132}{157}

	\section{Does the rat win?}

		We have arrived at the crux of the algorithm. Does the rat win for some integer $k$?

@@

		\begin{definition}\label{def:face-quiet}
			When the rat-catcher is in some face $f$, then edge $e$ is noisy iff. there is a closed walk of length scrictly less than $k$ containing $f^*$ and $e^*$ in the dual $M^*$.
		\end{definition}

@@

		\textbf{impl. carving width and the rat cathing algorithm}

		\ref{problem:cw}

@@

		\textbf{impl. noisy-subgraph}

@@

		No other path of the implementation needs the assumption that the dual is planar, therefore the output doesn't need to be a rotation system. This simplifies the implementation.

@@

@@

\newpage
\section{Appendix}
	% \pyfile{branch_decomposition.py}{0}{1000}
	% \pyfile{branch_width_brute_force.py}{0}{1000}	
	% \pyfile{branch_width.py}{0}{1000}
	% \pyfile{carving_decomposition.py}{0}{1000}
	% \pyfile{carving_width_brute_force.py}{0}{1000}
	% \pyfile{carving_width.py}{0}{1000}
	% \pyfile{contraction.py}{0}{1000}
	% \pyfile{dual_graph.py}{0}{1000}
	% \pyfile{Graph.py}{0}{1000}
	% \pyfile{medial_graph.py}{0}{1000}
	% \pyfile{parse_graph.py}{0}{1000}
	% \pyfile{parse_newick.py}{0}{1000}
